The role of cognateness in native spoken word recognition

### 2025-07-24 docs/manuscript.tex
Update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{float}
\usepackage{tabularray}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{rotating}
\UseTblrLibrary{siunitx}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\usepackage{tipa}
\usepackage{booktabs}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
\ifPDFTeX
\else
\babelfont{rm}[]{Latin Modern Roman}
\fi
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage[english]{selnolig} % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The role of cognateness in native spoken word recognition},
  pdfauthor={Gonzalo Garcia-Castro; Serene Siow; Kim Plunkett; Nuria Sebastian-Galles},
  pdflang={en},
  pdfkeywords={cognate, spoken word
recognition, phonology, bilingualism, non-native speech
processing, lexical access},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Cognateness and recognition of foreign
words }
\title{The role of cognateness in native spoken word
recognition\thanks{The authors declare no conflicts of interest with
regard to the funding source of this study. This study was supported by
the Spanish Ministry for Science and Innovation and State Research
Agency (Project PGC2018-101831-B-I00 financed by Spanish Ministry of
Science Innovation and Universities (MCIU), Spanish Research Agency
(AEI) and the European Regional Development Fund (ERDF), and by the
Economic and Social Research Council (ESRC) (ES/S010947/1, UK). GGC was
supported by a FPI research contract (PRE2019-088165). SS was supported
by XXXXXXXXXXXXX. KP was supported by XXXXXXXXXXXXXXXXX. NSG was
supported by an ICREA Academia 2019 award from the Catalan Institution
for Research and Advanced Studies (ICREA). We are grateful to Irina
Lepadatu and Nicola Gillen, and to the Speech Acquisition and Perception
research group for their helpful feedback. We thank Xavier Mayoral,
Silvia Blanch, Pamela Miller, and Cristina Cuadrado for their technical
support. We also thank all families and infants who participated in the
experiments. This study was conducted according to guidelines laid down
in the Declaration of Helsinki, and was approved by the Drug Research
Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference
2020/9080/I. The data and code necessary to reproduce the analyses
presented here are publicly accessible, as are the materials necessary
to attempt to replicate the findings. The analyses presented here were
not preregistered. Data, code, and materials for this research are
available at the following URLs: \url{https://osf.io/hy984/},
\url{https://github.com/gongcastro/translation-elicitation}.}}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Gonzalo
Garcia-Castro}~\orcidlink{0000-0002-8553-4209}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}\asep\textbf{Serene
Siow}~\orcidlink{0000-0001-6482-2191}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\href{mailto:siow.serene@gmail.com}{siow.serene@gmail.com}\asep\textbf{Kim
Plunkett}~\orcidlink{0000-0003-0216-7480}\\Department of Experimental
Psychology\\University of Oxford\\Oxford, OX2
6GG\\\href{mailto:kim.plunkett@psy.ox.ac.uk}{kim.plunkett@psy.ox.ac.uk}\asep\textbf{Nuria
Sebastian-Galles}~\orcidlink{0000-0001-6938-2498}\\Center for Brain and
Cognition\\Universitat Pompeu Fabra\\Barcelona,
08005\\\href{mailto:nuria.sebastian@upf.edu}{nuria.sebastian@upf.edu}}
\date{}

@@

\begin{abstract}
While listening to speech in an unfamiliar language, words and phrases
in the native language are activated, leading to common auditory
illusions like Soramimi, a particular case of homophonic translation.
Homophonic translation has received little attention as a
psycholinguistic phenomenon. In this study, we explored the mechanisms
underlying lexical activation and selection in homoponic translation.
Across three studies, English and Spanish native adults completed a
translation elicitation task, in which they listened to a series of
Catalan or Spanish words (to which participants were unfamiliar), and
then had to type their best-guess translation in their native language.
Both English and Spanish natives were surprisingly good at translating
unfamiliar words, efficiently exploiting the phonological similarity
between the presented (unfamiliar) words and their correct translation.
When the correct translation belonged to high-density phonological
neighbours, participants' ability to benefit from phonological
similarity decreased. Finally, Spanish natives exploited phonological
similarity more efficiently: provided identical phonological similarity,
Spanish natives were more likely to guess the correct translation of the
unfamiliar word than English natives. Overall, we show that speech in an
unfamiliar language triggers equivalent dynamics of lexical selection
than native speech, providing a psycholinguistic account for homophonic
translation.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
cognate \sep spoken word
recognition \sep phonology \sep bilingualism \sep non-native speech
processing \sep 
lexical access

@@

We further investigated the effect of phonological competition. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with other words
in the native language. Words with denser phonological neighbourhoods
are recognised more slowly and less accurately than words with sparser
phonological neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \&
Peereman, 2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al.,
1989}; \citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}; \citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
especially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
word (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). Such
interference effect of phonological neighbourhood density has also been
reported across languages (\citeproc{ref-weber2004lexical}{Weber \&
Cutler, 2004}). We hypothesised that the size of the facilitatory role
of phonological similarity would be inversely proportional to the amount
of higher-frequency phonological neighbours of the presented word in the
target language. In the case of non-native speech recognition, the
presence of cross-linguistic pairs which are phonologically similar but
differ in meaning (e.g., false friends) may act as distractors during
lexical access, obstructing the selection of the appropriate target
translation in the listener's lexicon. For instance, Otwinowska \&
Szewczyk (\citeproc{ref-otwinowska2019more}{2019}) reported that false
friends were disadvantaged relative to non-cognates by Polish second
language learners, in contrast to cognates which were known better. It
is therefore important to investigate the joint effect of both cognates
and false friends when investigating the effect that cross-linguistic
phonological similarity has on word recognition in a foreign language.
This could shed light on available strategies and challenges associated
with the early stages of second language acquisition. For instance, one
might expect English participants to incorrectly translate the Spanish
word \emph{botón} as \emph{bottom} instead of as its correct translation
\emph{button}, due to the combined effect of the close phonological
similarity between \emph{botón} and \emph{bottom} along with the high
lexical frequency of \emph{bottom} relative to \emph{button}. To test
this prediction, we developed a lexical frequency-dependent measure of
cross-language phonological neighbourhood density, in which a neighbour
is counted only if it is higher frequency and is one phoneme apart from
the presented word. If the phonological neighbourhood density of the
target translation affects participants' performance negatively, this
would suggest that competitors in the native language affect recognition
of non-native words during recognition of foreign speech. We conducted a
series of three experiments to test these predictions.

@@

All materials, data, and code used in this study are hosted in the Open
Science Framework
\href{https://osf.io/9fjxm/?view_only=aab7636ce1af48cf832596a7ea9101c5/}{https://osf.io/9fjxm/}
and a GitHub repository
\url{https://github.com/gongcastro/translation-elicitation.git}, along
with additional notes.

@@

\setlength{\LTpost}{0mm}

\begin{longtable}{l|rrrc}

\caption{\label{tbl-participants}Participant details.}

\tabularnewline

\toprule
\multicolumn{1}{l}{} &  & \multicolumn{2}{c}{Age} &  \\ 
\cmidrule(lr){3-4}
\multicolumn{1}{l}{} & N\textsuperscript{\textit{1}} & Mean ± SD  & Range & Second language \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 35 (8) & $21.80$ ± $2.08$ & $18$–$26$ & Russian (1) \\ 
cat-ENG & 36 (4) & $21.72$ ± $2.25$ & $18$–$25$ & French (1), German (1), Italian (1), Punjabi (1), Several (1) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & 33 (12) & $21.85$ ± $3.00$ & $18$–$33$ & French (9), German (1), Italian (2) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 32 (1) & $21.72$ ± $2.59$ & $18$–$26$ & German (2), Japanese (1) \\ 
cat-ENG & 32 (0) & $22.31$ ± $2.39$ & $18$–$26$ & Cantonese (1), Irish (1) \\ 
\bottomrule

\end{longtable}

\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}Number of included participants (number of excluded participants.)\\
\end{minipage}

@@

\paragraph{Data processing}\label{data-processing}

@@

\paragraph{Modelling approach and statistical
inference}\label{modelling-approach-and-statistical-inference}

@@

\begin{longtable}{l|rrrrrrrrr}

\caption{\label{tbl-dataset}Summary of participants' accuracy in the
translation elicitation task across Experiments 1, 2, and 3.}

\tabularnewline

>>>>

\documentclass[
  man,
  longtable,
  nolmodern,
  notxfonts,
  notimes,
  colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{apa7}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage[bidi=default]{babel}
\babelprovide[main,import]{english}

% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}

\RequirePackage{longtable}
\RequirePackage{threeparttablex}

\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
	{0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
	{-.5em}%
	{\normalfont\normalsize\bfseries\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{0.5em}%
	{0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
	{-\z@\relax}%
	{\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

\usepackage{longtable, booktabs, multirow, multicol, colortbl, hhline, caption, array, float, xpatch}
\usepackage{subcaption}

\renewcommand\thesubfigure{\Alph{subfigure}}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.7}

\usepackage{tcolorbox}
\tcbuselibrary{listings,theorems, breakable, skins}
\usepackage{fontawesome5}

\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{ACACAC}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582EC}
\definecolor{quarto-callout-important-color-frame}{HTML}{D9534F}
\definecolor{quarto-callout-warning-color-frame}{HTML}{F0AD4E}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02B875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{FD7E14}

%\newlength\Oldarrayrulewidth
%\newlength\Oldtabcolsep

\usepackage{hyperref}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{newtx}

\defaultfontfeatures{Scale=MatchLowercase}
\defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}

\title{The role of cognateness in native spoken word recognition}

\shorttitle{The role of cognateness in native spoken word recognition}

\usepackage{etoolbox}

\authorsnames[{1},{2},{2},{1}]{Gonzalo Garcia-Castro,Serene Siow,Kim
Plunkett,Nuria Sebastian-Galles}

\authorsaffiliations{
{Center for Brain and Cognition, Universitat Pompeu Fabra},{Department
of Experimental Psychology, University of Oxford}}

\leftheader{Garcia-Castro, Siow, Plunkett and Sebastian-Galles}

\abstract{When listening to speech in an unfamiliar language, words and
phrases in the native language are frequently activated due to their
acoustic similarity with some parts of the speech stream. This study
explored this phenomenon, for the first time, from a language processing
perspective. Across three studies, English- and Spanish-native adults
completed a translation elicitation task, in which they listened to a
series of words from an unfamiliar language (Catalan or Spanish for
English speakers, Catalan for Spanish speakers). For each presented
word, participants had to type their best-guess translation in their
native language. Both English and Spanish natives were surprisingly good
at translating unfamiliar words, efficiently exploiting the phonological
similarity between the presented (unfamiliar) words and their correct
translation. When the correct translation belonged to high-density
phonological neighbourhood, participants' ability to benefit from
phonological similarity decreased. Spanish participants, whose native
language was typologically closer to the presented language, benefited
more strongly from phonological similarity than English participants.
Overall, we show that speech in an unfamiliar language triggers
equivalent dynamics of lexical selection than native speech, providing a
psycholinguistic account for homophonic translation. }

\keywords{cognateness, spoken word recognition, phonology, speech
processing, non-native speech, lexical access}

\authornote{\par{\addORCIDlink{Gonzalo
Garcia-Castro}{0000-0002-8553-4209}}\par{\addORCIDlink{Serene
Siow}{0000-0001-6482-2191}}\par{\addORCIDlink{Kim
Plunkett}{0000-0003-0216-7480}}\par{\addORCIDlink{Nuria
Sebastian-Galles}{0000-0001-6938-2498}}

\par{       }
\par{Correspondence concerning this article should be addressed
to Gonzalo
Garcia-Castro, Email: \href{mailto:gonzalo.garciadecastro@upf.edu}{gonzalo.garciadecastro@upf.edu}}
}

\makeatletter
\let\endoldlt\endlongtable
\def\endlongtable{
\hline
\endoldlt
}
\makeatother
\RequirePackage{longtable}
\DeclareDelayedFloatFlavor{longtable}{table}

\urlstyle{same}

\usepackage{booktabs}

\usepackage{caption}

\usepackage{longtable}

\usepackage{colortbl}

\usepackage{array}

\usepackage{float}

\usepackage{tabularray}

\usepackage[normalem]{ulem}

\usepackage{graphicx}

\UseTblrLibrary{booktabs}

\UseTblrLibrary{rotating}

\UseTblrLibrary{siunitx}

\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}

\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}

\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}

\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{apxfig}{h}{loapxfig}}{\newfloat{apxfig}{h}{loapxfig}[chapter]}
\floatname{apxfig}{Appendix Fig.}
\newcommand*\listofapxfigs{\listof{apxfig}{List of Appendix Fig.s}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

% From https://tex.stackexchange.com/a/645996/211326
%%% apa7 doesn't want to add appendix section titles in the toc
%%% let's make it do it
\makeatletter
\xpatchcmd{\appendix}
  {\par}
  {\addcontentsline{toc}{section}{\@currentlabelname}\par}
  {}{}
\makeatother

%% Disable longtable counter
%% https://tex.stackexchange.com/a/248395/211326

\usepackage{etoolbox}

\makeatletter
\patchcmd{\LT@caption}
  {\bgroup}
  {\bgroup\global\LTpatch@captiontrue}
  {}{}
\patchcmd{\longtable}
  {\par}
  {\par\global\LTpatch@captionfalse}
  {}{}
\apptocmd{\endlongtable}
  {\ifLTpatch@caption\else\addtocounter{table}{-1}\fi}
  {}{}
\newif\ifLTpatch@caption
\makeatother

@@

\setlength\LTleft{0pt}

@@

We further investigated the effect of phonological competition. Even if
the presented word and its target translation share high phonological
similarity, participants may still provide incorrect translations if the
presented word also shares high phonological similarity with other words
in the native language. Words with denser phonological neighbourhoods
are recognised more slowly and less accurately than words with sparser
phonological neighbourhoods (\citeproc{ref-dufour2003lexical}{Dufour \&
Peereman, 2003}; \citeproc{ref-goldinger1989priming}{Goldinger et al.,
1989}; \citeproc{ref-hamburger1996phonological}{Hamburger \& Slowiaczek,
1996}; \citeproc{ref-luce1990similarity}{Luce et al., 1990};
\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). This is
especially true if such neighbours share higher phonological similarity
with the presented word, or are lexically more frequent than the target
word (\citeproc{ref-luce1998recognizing}{Luce \& Pisoni, 1998}). Such
interference effect of phonological neighbourhood density has also been
reported across languages (\citeproc{ref-weber2004lexical}{Weber \&
Cutler, 2004}). We hypothesised that the size of the facilitatory role
of phonological similarity would be inversely proportional to the amount
of higher-frequency phonological neighbours of the presented word in the
target language. In the case of non-native speech recognition, the
presence of cross-linguistic pairs which are phonologically similar but
differ in meaning (e.g., false friends) may act as distractors during
lexical access, obstructing the selection of the appropriate target
translation in the listener's lexicon. For instance, Otwinowska and
Szewczyk (\citeproc{ref-otwinowska2019more}{2019}) reported that false
friends were disadvantaged relative to non-cognates by Polish second
language learners, in contrast to cognates which were known better. It
is therefore important to investigate the joint effect of both cognates
and false friends when investigating the effect that cross-linguistic
phonological similarity has on word recognition in a foreign language.
This could shed light on available strategies and challenges associated
with the early stages of second language acquisition. For instance, one
might expect English participants to incorrectly translate the Spanish
word \emph{botón} as \emph{bottom} instead of as its correct translation
\emph{button}, due to the combined effect of the close phonological
similarity between \emph{botón} and \emph{bottom} along with the high
lexical frequency of \emph{bottom} relative to \emph{button}. To test
this prediction, we developed a lexical frequency-dependent measure of
cross-language phonological neighbourhood density, in which a neighbour
is counted only if it is higher frequency and is one phoneme apart from
the presented word. If the phonological neighbourhood density of the
target translation affects participants' performance negatively, this
would suggest that competitors in the native language affect recognition
of non-native words during recognition of foreign speech. We conducted a
series of three experiments to test these predictions.

@@

All materials, data, and code used in this study are hosted in the Open
Science Framework
\href{https://osf.io/9fjxm/?view_only=aab7636ce1af48cf832596a7ea9101c5/}{https://osf.io/9fjxm/}
and a GitHub repository
\url{https://github.com/gongcastro/translation-elicitation.git}, along
with additional notes. For reproducibility, a Docker image of the
RStudio session is available on DockerHub
(\url{https://hub.docker.com/repository/docker/gongcastro/translation-elicitation/general}).

@@

{\caption{{Participant details.}{\label{tbl-participants}}}
\vspace{-20pt}}

\setlength{\LTpost}{0mm}
\begin{longtable*}{l|rrrc}
\toprule
\multicolumn{1}{l}{} &  & \multicolumn{2}{c}{Age} &  \\ 
\cmidrule(lr){3-4}
\multicolumn{1}{l}{} & N\textsuperscript{\textit{1}} & Mean ± SD  & Range & Second language \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 35 (8) & $21.80$ ± $2.08$ & $18$–$26$ & Russian (1) \\ 
cat-ENG & 36 (4) & $21.72$ ± $2.25$ & $18$–$25$ & French (1), German (1), Italian (1), Punjabi (1), Several (1) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & 33 (12) & $21.85$ ± $3.00$ & $18$–$33$ & French (9), German (1), Italian (2) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 32 (1) & $21.72$ ± $2.59$ & $18$–$26$ & German (2), Japanese (1) \\ 
cat-ENG & 32 (0) & $22.31$ ± $2.39$ & $18$–$26$ & Cantonese (1), Irish (1) \\ 
\bottomrule
\end{longtable*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}Number of included participants (number of excluded participants.)\\
\end{minipage}

\end{table}

@@

\paragraph{Data processing.}\label{data-processing}

@@

\paragraph{Modelling approach and statistical
inference.}\label{modelling-approach-and-statistical-inference}

@@

{\caption{{Summary of participants' accuracy in the translation
elicitation task across Experiments 1, 2, and 3.}{\label{tbl-dataset}}}
\vspace{-20pt}}
